---
title: "Lab Meeting Demo -- Data Visualization: `ggplot` part 4"
output:
  html_document:
    theme: spacelab
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

## Lab meeting goals
* solicit feed back & ideas on how to introduce these concepts to people with non-begginer familiarity with the topic
* solicit content ideas and general knowledge
  + what r markdown features do you know and love?
  + if you recall, what are things you wish you'd learned about earlier?

## Learning Goals
*At the end of this exercise, you will be able to:*    
1. Create Violin plots.
2. Create Violin plots.Build heat maps using `ggplot.` 
3. Customize labels on axes using `labs` and `themes`.  
4. Use `color`, `fill`, and `group` to customize plots and improve overall aesthetics.  


## Resources
- [ggplot2 cheatsheet](https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf)
- [R for Data Science](https://r4ds.had.co.nz/)
- [R Cookbook](http://www.cookbook-r.com/)
- [`ggplot` themes](https://ggplot2.tidyverse.org/reference/ggtheme.html)
- [Rebecca Barter `ggplot` Tutorial](http://www.rebeccabarter.com/blog/2017-11-17-ggplot2_tutorial/)
- [R Color Brewer](http://colorbrewer2.org/#type=sequential&scheme=BuGn&n=3)

## Install `RColorBrewer`, `paletteer`, and `ggthemes`
`RColorBrewer` and `paletteer` are helpful in selecting palettes of color that work with your data. 
```{r, eval=FALSE, echo=TRUE}
#install.packages("RColorBrewer")
#install.packages("paletteer")
#install.packages("ggthemes")
```

## Load the libraries
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(RColorBrewer)
library(paletteer)
library(janitor)
library(here)
```

```{r}
options(scipen=999) #cancels the use of scientific notation for the session
```

##Resources
- [ggplot2 cheatsheet](https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf)
- [`ggplot` themes](https://ggplot2.tidyverse.org/reference/ggtheme.html)
- [Rebecca Barter `ggplot` Tutorial](http://www.rebeccabarter.com/blog/2017-11-17-ggplot2_tutorial/)

## Load the libraries
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(here)
library(janitor)
```

## Load the data

Let's revisit the mammal life history data to practice our ggplot skills. The [data](http://esapubs.org/archive/ecol/E084/093/) are from: *S. K. Morgan Ernest. 2003. Life history characteristics of placental non-volant mammals. Ecology 84:3402.*

```{r message=FALSE, warning=FALSE}
life_history <- read_csv(here("data", "mammal_lifehistories_v2.csv")) %>% na_if("-999") %>% clean_names()
```


## Violin Plots
Violin plots show the distribution of continuous data across a categorical variable.
Violin plots are similar to box plots and dot plots

lets get an idea of the data structure
```{r}
life_history %>% 
  count(order, sort=T)
```
```{r}
str(life_history)
```

Here, we visualize the distribution of average mass across orders. 
Since the orders are so diverse in size (an ant v.s. a giraffe), lets use a log function on mass. 
Also, let's flip the axes to make the graph more friendly for landscape orientation.
```{r}
life_history %>%
  ggplot(aes(x = order, y = log(mass))) +
  geom_boxplot(na.rm = TRUE)+
  coord_flip()
```


Let's focus in on a small set of primates, _Cercopithecidae_ family, i.e. Old-world monkeys, and generate a boxplot for each taxonomic genus.
```{r}
life_history %>%
  filter(family == "Cercopithecidae")%>%
  ggplot(aes(x = genus, y = mass)) +
  geom_boxplot(na.rm = TRUE) +
  coord_flip()
```

Box plots are great at showing median, skew, and outliers. 
We can visualize skew better by adding a point layer with `geom_point()`

```{r}
life_history %>%
    filter(family == "Cercopithecidae")%>%
    ggplot(aes(x = genus, y = mass)) +
    geom_boxplot(color = "grey", na.rm = TRUE) +
    geom_point(alpha = 0.5, color = "red", na.rm = TRUE) + #alpha indicates transparency 
    coord_flip()
```
By using the geom_point() layer, we see that there are only a few observations for each group (genus).
Lets further zoom in on 3 genera of monkey: Macaca, Cercopithecus, and Trachypithecus


we can use violin plots to visualize the distribution of these points on another axis:

```{r, warning=FALSE}
life_history %>%
  filter(genus == "Macaca" | genus == "Cercopithecus"| genus == "Trachypithecus")%>%
  ggplot(aes(x = genus, y = log(mass))) +
  geom_boxplot(color = "grey", na.rm = TRUE) +
  geom_point(alpha = 0.3, color = "red", na.rm = TRUE) + #alpha indicates transparency 
  geom_violin(alpha = .3, na.rm = TRUE) +
  coord_flip()
```



Lets zoom out and apply this graph style to a few primate families, Cercopithecidae, Hominidae and Cebidae 
```{r}
life_history %>%
  filter(family == "Cercopithecidae" | family == "Indridae" | family == "Cebidae") %>%
  ggplot(aes(x = family, y = mass)) +
  geom_point(alpha = 0.3, color = "red", na.rm = TRUE) + #alpha indicates transparency 
  geom_violin(alpha = .3, na.rm = TRUE) +
  coord_flip()
```
we can use violin plots to visualize the distribution of these points on another axis:

by default, the violin plot height is based on normalized area. we can switch to count with:

```{r}
life_history %>%
  filter(family == "Cercopithecidae" | family == "Indridae" | family == "Cebidae") %>%
  ggplot(aes(x = family, y = mass)) +
  geom_point(alpha = 0.3, color = "red", na.rm = TRUE) + #alpha indicates transparency 
  geom_violin(scale = "count", alpha = .3, na.rm = TRUE) +
  coord_flip()
```



## Heat maps
[heat maps](https://www.r-graph-gallery.com/79-levelplot-with-ggplot2.html) are a great tool for visualizing data across 2 categorical variables.


## Example 1: column names are data
The following data show results from a drug trial with four treatments on six patients. The values represent resting heart rate.  
```{r}
heartrate <- readr::read_csv("data/heartrate.csv")
heartrate
```

Let's pivot this data, so there is no data stored in variables
```{r}
heartrate_long <- heartrate %>% 
  pivot_longer(-patient, #patient does not pivot
               names_to = "drug", 
               values_to = "heartrate"
               )
heartrate_long
```

Now we can use `geom_tile()` to represent the contents
```{r}
heartrate_long %>%
  ggplot(aes(patient, drug, fill= heartrate)) + 
  geom_tile() +
  scale_fill_gradient(low="yellow", high="red")
```



Lets try this on a larger data set. 
Remember this one from the [United Nations Food and Agriculture Organization](http://www.fao.org/about/en/) on world fisheries? These data are pretty wild, so we need to do some cleaning. First, load the data.  

Load the data `FAO_1950to2012_111914.csv` as a new object titled `fisheries`.
```{r}
fisheries <- readr::read_csv(file = "data/FAO_1950to2012_111914.csv")
```



clean up names, 
```{r}
fisheries <- janitor::clean_names(fisheries)
names(fisheries)
```

lets practice using across() to change some interesting variables to factors
```{r}
fisheries <- 
  fisheries %>% mutate(across(c(country, isscaap_group_number, asfis_species_number, fao_major_fishing_area), as_factor))
```

Notice that these data are in wide format - year (a variable) is represented as columns. Let's practice pivoting long so we can make a heatmap: 
```{r}
fisheries_long <- fisheries %>% 
pivot_longer(cols = starts_with("x"),
               names_to = "year",
               values_to = "catch",
               values_drop_na = TRUE) %>% 
  mutate(year= as.numeric(str_replace(year, 'x', ''))) %>% 
  mutate(catch= str_replace(catch, c(' F'), replacement = '')) %>% 
  mutate(catch= str_replace(catch, c('...'), replacement = '')) %>% 
  mutate(catch= str_replace(catch, c('-'), replacement = '')) %>% 
  mutate(catch= str_replace(catch, c('0 0'), replacement = ''))

fisheries_long$catch <- as.numeric(fisheries_long$catch)
```


which countries caught a lot of squid in the 90's?
```{r}

fisheries_long %>%  
  filter(isscaap_taxonomic_group == "Squids, cuttlefishes, octopuses") %>% 
  filter(between(year, 1990, 1999)) %>%
  group_by(country) %>% 
  summarize(catch_total=sum(catch, na.rm=T)) %>% 
  arrange(desc(catch_total))
```

```{r}

fish_tycoons <-   fisheries_long%>%
  filter(isscaap_taxonomic_group == "Squids, cuttlefishes, octopuses") %>% 
  filter(between(year, 1990, 1999)) %>%
   filter( country == "Japan" |			
   country == "Korea, Republic of" |
   country == "Taiwan Province of China" |
  country == "Argentina" |
   country == "China" |
   country == "Mexico"	|
   country == "Thailand" |
   country == "France" |
   country == "United States of America" |
   country == "Spain"
            )%>%
  group_by(country) %>% 
  summarize(catch_total=sum(catch, na.rm=T), catch = catch, year = year) %>% 
  arrange(desc(catch_total))
  
```
```{r}
fish_tycoons%>%
  ggplot(aes(x = year, y = country, fill= log(catch))) + 
  geom_tile(na.rm = T) +
  scale_fill_gradient(low="yellow", high="red")

```


## Formatting reports: recapping markdown

check out this (source)[https://rmarkdown.rstudio.com/authoring_basics.html] for more

We know about code chunks

```{r}
names(heartrate_long)
```

but we can also use back ticks to put a small bit of code in `line with the text`

we can insert hyperlinks into out reports:
like (this)[https://en.wikipedia.org/wiki/Hyperlink]



